---
title: "bayesBirdModel_multiSPecies_v10 - FARM"
author: "Liz Law"
date: "`r Sys.Date()`"
output: html_document
---
workingconservation@gmail.com

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Multi-species occupancy model using nimbleEcology
* marginalized occupancy likelihood (docc)
* uniform priors
* variables are normalised prior to analysis
* community occupancy hyperparameter sd.betalpsi not parameter specific
* community detection hyperparameter mu.betalp and sd.betalp not parameter specific
* rare species share lpsi, lp, and beta params

# Setup
```{r library setup}
library(nimble)
library(nimbleEcology)
library(tidyverse)

mySeed <- 101 
set.seed(mySeed)

wd <- "/Users/elaw/Desktop/LeuphanaProject/BirdModelling/ETH_birds"
data_input <- paste0(wd, "/Data/nimbleModel_multiOcc_v6_allData.RDS")
#wd <- "C:/Users/chloe.nater/OneDrive - NINA/Documents/Projects/Other/BirdOccupancy_Liz"
#data_input <- paste0(wd, "/Data/nimbleModel_multiOcc_v6_allData.RDS")
results_folder <- paste0(wd, "/Results/")  
version_folder <- "v10/"
```

# build model
```{r model code}
occupancy_code <- nimbleCode({
  
  # Priors
  
  # Hyperpriors describe the community
  # To constrain the model as much as is reasonable, we used constant sd for 
  # all beta psi (occupancy) parameters, and a constant mu and sd for all beta p 
  # (observation) parameters.
  
  # For the model of occupancy (psi)
    Mu.lpsi ~ dunif(0,1)          # uniform community prior  
    mu.lpsi <- logit(Mu.lpsi)     # logit model link
    sd.lpsi ~ dunif(0, 12)        # bounds of uniform chosen by trial and error
         
    for (lpsii in 1:npsi){
      mu.betalpsi[lpsii] ~ dunif(-5, 5) # mu for each psi param drawn from uniform prior 
    }
    sd.betalpsi  ~ dunif(0, 6) # constant sd for all species on all psi parameters
  
  # For the model of detection (p)
    Mu.lp ~ dunif(0,1)          # uniform community prior
    mu.lp <- logit(Mu.lp)       # logit model link
    sd.lp ~ dunif(0, 5)         # bounds of uniform chosen by trial and error
    
    for (lpi in 1:np){
      mu.betalp[lpi] ~ dunif(-5, 5) # mu for each p param drawn from uniform prior 
    }
    sd.betalp ~ dunif(0, 6) # constant sd for all species on all psi parameters
  
  # Priors for species-specific effects in occupancy and detection
  # are derived from the community level distributions
  # Here, we constrain the model further, by specifying that rare species share
  # parameters. We have ordered the species by number of observations, and 
  # defined rare and common species.
  
  ## COMMON SPECIES
  for(k in 1:commonSpp){   
    # occupancy 
    lpsi[k] ~ dnorm(mu.lpsi, sd=sd.lpsi)    
    for (lpsii in 1:npsi){
      betalpsi[k, lpsii] ~ dnorm(mu.betalpsi[lpsii], sd=sd.betalpsi)
    }
    # detection
    lp[k] ~ dnorm(mu.lp, sd=sd.lp)
  }       
  
  ## RARE SPECIES - share params
  ### first make the rare params
  lpsiRARE ~ dnorm(mu.lpsi, sd=sd.lpsi) 
  for (lpsii in 1:npsi){
    betalpsiRARE[lpsii] ~ dnorm(mu.betalpsi[lpsii], sd=sd.betalpsi)
  }
  lpRARE ~ dnorm(mu.lp, sd=sd.lp)
  
  ### then insert it into the matrix 
  for(k in rareSpp:M){
    # occupancy
    lpsi[k] <- lpsiRARE
    for (lpsii in 1:npsi){
      betalpsi[k, lpsii] <- betalpsiRARE[lpsii]
    }
    
    # detection 
    lp[k] <- lpRARE
  }
  
  # detection, variable coefficient for each variable, but non-species specific
  for(lpi in 1:np){
    betalp[lpi] ~ dnorm(mu.betalp[lpi], sd=sd.betalp)
  }
  
  # the multi-species occupancy model (for each species in each site)
  
  for(k in 1:M){
    for (i in 1:nSite){
      
      logit(psi[i,k]) <- lpsi[k] + 
        betalpsi[k,1] * Xoc[i,1] +  # elevation
        betalpsi[k,2] * Xoc[i,2] +  # fl_dis
        betalpsi[k,3] * Xoc[i,3] +  # sidi1ha
        betalpsi[k,4] * Xoc[i,4] +  # slope
        betalpsi[k,5] * Xoc[i,5]    # farm_type
      
      # mu.psi[i,k] <- psi[i,k]       
      # z[i,k] ~ dbern(mu.psi[i,k])       ## occupancy drawn from psi

      for(j in 1:nVisits){
        
        logit(p[i,j,k]) <- lp[k] + 
          betalp[1] * Xob[i,j,1] + # date
          betalp[2] * Xob[i,j,2] + # start1
          betalp[3] * Xob[i,j,3] + # start2
          betalp[4] * Xob[i,j,4]   # visibility
        
        # mu.p[i,j,k] <- z[i,k] * p[i,j,k]   
        # Y[i,j,k] ~ dbern(mu.p[i,j,k])     ## observations from occupancy and detection (p)

      }
      # Marginalized likelihood (dOcc_v from nimbleEcol)
       Y[i,1:nVisits,k] ~ dOcc_v(
        probOcc = psi[i,k], 
        probDet = p[i,1:nVisits,k], 
        len = nVisits)
    }
  }
})
```

# Specify model: code, data (observed), constants (set), and inits (starting value for estimated values)

```{r import data}
nimData <- readRDS(data_input)

y_all <- nimData$y_all
farmSites <- nimData$farmSites
OccVarsFarm <- nimData$OccVarsFarm
ObsVarsFarm <- nimData$ObsVarsFarm
sppTraits <- nimData$sppTraits
Nadd <- 17 # based on individual rarefaction curves for total farmland observations

y <- y_all[(dimnames(y_all)[[1]] %in% farmSites), , ]  # [site, visit, species]

nSite <- dim(y)[1]
nVisits <- dim(y)[2]
Nobs <- dim(y)[3]
M <- Nobs + Nadd
Y <- abind::abind(y, array(0, dim = c(dim(y)[1:2], Nadd)))  ## add the 'missing' species - not seen so = 0, 'absence'

Xoc <- OccVarsFarm %>% select(-point_id) %>% as.matrix() %>% array(dim = dim(.))   # [site, vars]
Xob <- array(NA, dim = c(nSite, nVisits, dim(ObsVarsFarm)[2]-2)) # [site, visit, vars]
Xob[,,1] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(date)
Xob[,,2] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(start1)
Xob[,,3] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(start2)
Xob[,,4] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(visibility)

# constrain to selected variables
Xoc <- Xoc[,1:5]  # elevation fl_dis sidi1ha   slope farm_type
Xob <- Xob[,,1:4] 

npsi <- dim(Xoc)[2]
np <- dim(Xob)[3]

# groups of interest
fspp <- sppTraits$forSpec 
mig <- sppTraits$migYes 
fnDiet <- sppTraits$FruiNect 
invDiet <- sppTraits$Invertebrate 

# common/rare vector numbers
rareSpp <- min(which(sppTraits$sumObs1))
commonSpp <- rareSpp-1

data <- list(Y = Y, 
             Xoc = Xoc,
             Xob = Xob)  
constants <- list(nSite = nSite, 
                  nVisits = nVisits, 
                  Nobs = Nobs,
                  Nadd = Nadd, 
                  M = M,
                  npsi = npsi,
                  np = np,
                  rareSpp=rareSpp,
                  commonSpp=commonSpp,
                  fspp=fspp,
                  mig=mig,
                  fnDiet=fnDiet,
                  invDiet=invDiet)

# Initial values - match up with the number of observed vars used in the model
lpsi <- rnorm(n = M)
betalpsi <- array(rnorm(n = M * npsi), dim = c(M, npsi))
lp <- rnorm(n = M)
betalp <- rnorm(n =  np) 

inits <- list(
  # z = zst, 
  lpsi = lpsi, 
  betalpsi = betalpsi, 
  lp = rnorm(n = M), 
  betalp = betalp 
)

# Set 1
params1 <- c("lpsi", "betalpsi", "lp", "betalp", "psi", "p")

```

## run MCMC
```{r run nimbleMCMC}

# MCMC settings - development - final
# ni <- 10      ;   nt <- 1    ;   nb <- 0       ;   nc <- 3
# ni <- 2500   ;   nt <- 5   ;   nb <- 0   ;   nc <- 3
ni <- 30000   ;   nt <- 5   ;   nb <- 10000   ;   nc <- 3
# (ni - nb)/nt # samples per chain = 4000 in the final models

samples <- nimbleMCMC(
  code = occupancy_code,
  data = data,
  constants = constants,
  inits = inits,
  monitors = params1,
  thin = nt,
  niter = ni,
  nburnin = nb,
  nchains = nc)
```

# model diagnostics
```{r MCMCvis}

library(MCMCvis)
pg1 <- c("lpsi", "betalpsi", "lp", "betalp") 

# summary of estimates --- we want Rhat < 1.1, and n.eff >100
mysum <- MCMCsummary(samples, 
            params = pg1, 
            round = 2)

mysum[which(mysum$Rhat>1.1),]
mysum[which(mysum$n.eff<100),]  

# trace plots
MCMCtrace(samples, 
          params = pg1,
          ind = TRUE, 
          Rhat = TRUE, n.eff = TRUE,
          filename = paste0("Results/", version_folder, "plots_traceDensity_pg1_farm_", format(Sys.time(), "%Y%m%d"), ".pdf"))

# caterpillar plots - choose parameter option
MCMCplot(samples, 
         #params = "lp",
         #params = "betalp",
         #params = "lpsi",
         #params = 'betalpsi\\[\\d{1,3}[,][ ][1]\\]', ISB = FALSE, exact = FALSE,  
         #params = 'betalpsi\\[\\d{1,3}[,][ ][2]\\]', ISB = FALSE, exact = FALSE, 
         #params = 'betalpsi\\[\\d{1,3}[,][ ][3]\\]', ISB = FALSE, exact = FALSE, 
         #params = 'betalpsi\\[\\d{1,3}[,][ ][4]\\]', ISB = FALSE, exact = FALSE, 
         #params = 'betalpsi\\[\\d{1,3}[,][ ][5]\\]', ISB = FALSE, exact = FALSE, 
         #params = "psi", 
         #params = "p",
         ci = c(50, 90), 
         HPD = FALSE)  # plots equal tailed credible intervals
```


```{r save results and data inputs}
saveRDS(samples, paste0(results_folder, version_folder, "BirdOccMod_dOcc_samples_farm_", format(Sys.time(), "%Y%m%d"), "_seed", mySeed, ".RDS"))
saveRDS(list(data,constants), paste0("Data/Farm_nimbleOccData_v6_", format(Sys.time(), "%Y%m%d"), ".RDS"))
```
