---
title: "Examine FARM models and their projections v9"
output: html_document
---
workingconservation@gmail.com

# Setup libraries
```{r setup 1, message=FALSE}

library(tidyverse)
library(terra)
library(MCMCvis)

# ggplot theme set
theme_set(theme_bw())
```

# setup data
```{r setup 2}
wd <- "/Users/elaw/Desktop/LeuphanaProject/BirdModelling/ETH_birds"
results_folder <- paste0(wd, "/Results/")  
version_folder <- "v9/"
mydate <- 20221017
samples <- readRDS(paste0(results_folder, version_folder, "BirdOccMod_dOcc_samples_farm_", mydate, ".RDS"))
data_input <- readRDS(paste0(wd, "/Data/Farm_nimbleOccData_v6_", mydate, ".RDS"))
list2env(data_input[[1]], envir = .GlobalEnv); list2env(data_input[[2]], envir = .GlobalEnv); rm(data_input)

rast_folder <- "/Users/elaw/Desktop/LeuphanaProject/ETH_SpatialData/data_10m/Baseline/"
farm_stack <- rast(paste0(rast_folder, "farm_stack_10m.tif"))
pred_stack <- rast(paste0(results_folder, version_folder, "farm_pred_stack.tif"))

rast_vals <- values(pred_stack, na.rm=TRUE) %>% as_tibble()
site_vals <- as_tibble(Xoc); names(site_vals) <- names(rast_vals)
```

# site vars vs raster vars
```{r summarise vars}
summarise_all(rast_vals, range) %>% t()
summarise_all(site_vals, range) %>% t()
```

# Plot histograms
Site sampled variables (green) are typically a biased sub-section of the range of variables in the rasters (purple). This is not as much of an issue as in forest.
```{r var histograms}
plot_hists <- function(myvar, binwidth=0.25){
  mc <- viridis::viridis(3)
  ggplot(rast_vals, aes(x={{myvar}}, y = stat(density))) +
    geom_histogram(binwidth = binwidth, fill=mc[1], alpha = 0.5) + 
    geom_histogram(data = site_vals, binwidth = binwidth, fill=mc[2], alpha = 0.5)
}

plot_hists(elevation)
plot_hists(fldis)
plot_hists(sidi1ha)
```

# Examine betalpsi parameters from each species 

```{r species betas}
# select a reasonable set of samples 
mydraw <- 501:1500
mysamples <- list(
  chain1 = samples$chain1[mydraw, ],
  chain2 = samples$chain2[mydraw, ],
  chain3 = samples$chain3[mydraw, ]
)

# extract betas
b0 <- MCMCsummary(mysamples, 
                  params = 'lpsi\\[\\d{1,3}\\]', ISB = FALSE, exact = FALSE,  
                  round = 2)
b1 <- MCMCsummary(mysamples, 
                  params = 'betalpsi\\[\\d{1,3}[,][ ][1]\\]', ISB = FALSE, exact = FALSE,  
                  round = 2)
b2 <- MCMCsummary(mysamples, 
                  params = 'betalpsi\\[\\d{1,3}[,][ ][2]\\]', ISB = FALSE, exact = FALSE,  
                  round = 2)
b3 <- MCMCsummary(mysamples, 
                  params = 'betalpsi\\[\\d{1,3}[,][ ][3]\\]', ISB = FALSE, exact = FALSE,  
                  round = 2)

tibble(
  ElevationMean = summary(b1$mean > 0), ElevationMedian = summary(b1$`50%`>0),
  FarmDistanceMean = summary(b2$mean > 0), FarmDistanceMedian = summary(b2$`50%`>0),
  Sidi1haMean = summary(b3$mean > 0), Sidi1haMedian = summary(b3$`50%`>0)
) %>% t %>% 
  as_tibble(rownames = "SummaryType") %>% 
  select(-V1, Negative=V2, Positive=V3)
  
```

Plot the betalpsi params 
```{r plot beta catterpillars}
plotbeta <- function(bp, bname, fgroup=NULL, fgroupName){
  bp <- bind_cols(bp, 
                  fspp=c(fspp, rep("unknown", Nadd)), 
                  mig=c(mig, rep("unknown", Nadd)),
                  fnDiet=c(fnDiet, rep("unknown", Nadd)),
                  invDiet=c(invDiet, rep("unknown", Nadd)))
  ggplot(arrange(bp,`50%`), 
         aes(x=`50%`, xmin=`2.5%`, xmax=`97.5%`,
             y=1:M))+
    geom_linerange(aes(color={{fgroup}}))+
    geom_point(aes(x=mean), color="darkgrey")+
    geom_point(aes(color={{fgroup}}))+
    geom_vline(xintercept=0, color="red")+
    scale_color_viridis_d()+
    labs(x=paste0(bname," beta parameter:\n95% HDI (lines), mean (grey points), median (color points)"),
         y="Species, sorted by median beta",
         color = fgroupName)+
    theme(axis.text.y=element_blank(),
          axis.ticks.y=element_blank())
}

plotbeta(b1, "Elevation", fspp, "Forest species")
plotbeta(b2, "Farm distance", fspp, "Forest species")
plotbeta(b3, "Landscape diversity (sidi1ha)", fspp, "Forest species")

plotbeta(b1, "Elevation", mig, "Migratory species")
plotbeta(b2, "Farm distance", mig, "Migratory species")
plotbeta(b3, "Landscape diversity (sidi1ha)", mig, "Migratory species")
```

Plot beta relationships
```{r plot beta relationships}
newdata <- tibble(
  elevation = seq(-2.811507, 3.647551, length.out=100) %>% rep(2),
  mElevation = mean(rast_vals$elevation),
  fldis = seq(-2.126014, 3.944279, length.out=100) %>% rep(2),
  mFldis = mean(rast_vals$fldis),
  sidi1ha = seq(-1.156326, 3.160600, length.out=100) %>% rep(2),
  mSidi1ha = mean(rast_vals$sidi1ha)
)

elev_mfd_mld_psi <- matrix(data=NA, nrow = nrow(newdata), ncol = M)
for(k in 1:M){
  logitpsi <- b0$mean[k]  +
    b1$mean[k] * newdata$elevation + 
    b2$mean[k] * newdata$mFldis +
    b3$mean[k] * newdata$mSidi1ha 
  
  elev_mfd_mld_psi[,k] <- exp(logitpsi)/(exp(logitpsi) + 1)
}

melev_fldis_mld_psi <- matrix(data=NA, nrow = nrow(newdata), ncol = M)
for(k in 1:M){
  logitpsi <- b0$mean[k]  +
    b1$mean[k] * newdata$mElevation + 
    b2$mean[k] * newdata$fldis +
    b3$mean[k] * newdata$mSidi1ha 
  
  melev_fldis_mld_psi[,k] <- exp(logitpsi)/(exp(logitpsi) + 1)
}

melev_mfd_sidi_psi <- matrix(data=NA, nrow = nrow(newdata), ncol = M)
for(k in 1:M){
  logitpsi <- b0$mean[k]  +
    b1$mean[k] * newdata$mElevation + 
    b2$mean[k] * newdata$mFldis +
    b3$mean[k] * newdata$sidi1ha 
  
  melev_mfd_sidi_psi[,k] <- exp(logitpsi)/(exp(logitpsi) + 1)
}


out_psi <- elev_mfd_mld_psi %>% 
  as_tibble() %>% 
  bind_cols(newdata) %>% 
  pivot_longer(cols = V1:V176, names_to = "species", values_to = "elev_mfd_mld_psi") %>% 
  mutate(species = str_replace(species, "V", "sp")) %>% 
  bind_cols(
    melev_fldis_mld_psi %>% 
      as_tibble() %>% 
      pivot_longer(cols = V1:V176, names_to = "species", values_to = "melev_fldis_mld_psi") %>% 
      select(-species)
  ) %>% 
    bind_cols(
    melev_mfd_sidi_psi %>% 
      as_tibble() %>% 
      pivot_longer(cols = V1:V176, names_to = "species", values_to = "melev_mfd_sidi_psi") %>% 
      select(-species)
  ) %>% 
  add_column(
    b1 = rep(b1$mean, 200),
    b2 = rep(b2$mean, 200),
    b3 = rep(b3$mean, 200),
    fspp = rep(c(fspp, rep(FALSE, Nadd)),200), 
    fnDiet=rep(c(fnDiet, rep(FALSE, Nadd)),200), 
    invDiet=rep(c(invDiet, rep(FALSE, Nadd)),200),
    observed=rep(c(rep("observed", Nobs), rep("absent", Nadd)), 200)
  ) %>% 
  mutate(fspp = if_else(fspp==TRUE, "fspp", "other"),
         fnDiet = if_else(fnDiet==TRUE, "fnDiet", "other"),
         invDiet = if_else(invDiet==TRUE, "invDiet", "other"))

plotpsi <- function(xvar, yvar, xname, yname, oxmin=-Inf, oxmax=Inf){
  ggplot(out_psi, 
         aes(x={{xvar}}, y={{yvar}}, color=species)) +
    # add rectangles showing non-sampled projection area
    annotate("rect", xmin = -Inf, xmax = oxmin, ymin = -Inf, ymax = Inf, 
             alpha =0.3)+
    annotate("rect", xmin = oxmax, xmax = Inf, ymin = -Inf, ymax = Inf, 
             alpha =0.3)+
    geom_line(alpha=0.5) + 
    scale_color_viridis_d()+
    theme(legend.position = "none") +
    labs(x=xname, y=yname)
}

plotpsi(elevation, elev_mfd_mld_psi, 
        "Elevation (center scaled, Farm distance and sidi1ha at mean)", "psi", 
        -1.788999, 1.629391)

plotpsi(fldis, melev_fldis_mld_psi, 
        "Farm distance (center scaled, Elevation and sidi1ha at mean)", "psi", 
        -1.380924, 2.606278)

plotpsi(sidi1ha, melev_mfd_sidi_psi, 
        "Landscape diversity (sidi1ha center scaled, Elevation and farm distance at mean)", "psi", 
        -1.156326, 2.282887)

```
Sum species over the parameters
```{r plot SR relationships}

plotSR <- function(xvar, y, xname, yname, oxmin=-Inf, oxmax=Inf){
  bind_cols(
    newdata,
    SR = rowSums(y),
    SRfspp = rowSums(y[,fspp]),
    SRmig = rowSums(y[,mig]),
    SRfnDiet = rowSums(y[,fnDiet]),
    SRinvDiet = rowSums(y[,invDiet])
  ) %>% 
    pivot_longer(cols = SR:SRinvDiet, names_to = "SpeciesSelection", values_to = "SR") %>% 
  ggplot(., 
         aes(x={{xvar}}, y=SR, color=SpeciesSelection), lwd=2) +
    # add rectangles showing non-sampled projection area
    annotate("rect", xmin = -Inf, xmax = oxmin, ymin = -Inf, ymax = Inf, 
             alpha =0.3)+
    annotate("rect", xmin = oxmax, xmax = Inf, ymin = -Inf, ymax = Inf, 
             alpha =0.3)+
    geom_line() + 
    scale_color_viridis_d()+
    labs(x=xname, y=yname)
}

plotSR(elevation, elev_mfd_mld_psi, 
        "Elevation (center scaled, Farm distance and sidi1ha at mean)", "Species Richness (sum PSI)", 
        -1.788999, 1.629391)

plotSR(fldis, melev_fldis_mld_psi, 
        "Farm distance (center scaled, Elevation and sidi1ha at mean)", "Species Richness (sum PSI)", 
        -1.380924, 2.606278)

plotSR(sidi1ha, melev_mfd_sidi_psi, 
        "Landscape diversity (sidi1ha center scaled, Elevation and farm distance at mean)", "Species Richness (sum PSI)", 
        -1.156326, 2.282887)

```

# Probability of detection 
```{r}

MCMCsummary(mysamples, 
                  params = "betalp",  
                  round = 2)

# extract betas for detection
d0 <- MCMCsummary(mysamples, 
                  params = "lp",  
                  round = 2)
d1 <- MCMCsummary(mysamples, 
                  params = 'betalp\\[[1]\\]', ISB = FALSE, exact = FALSE, 
                  round = 2)
d2 <- MCMCsummary(mysamples, 
                  params = 'betalp\\[[2]\\]', ISB = FALSE, exact = FALSE,  
                  round = 2)
d3 <- MCMCsummary(mysamples, 
                  params = 'betalp\\[[3]\\]', ISB = FALSE, exact = FALSE,  
                  round = 2)

logitp_v1 <- matrix(d0$mean, length(d0$mean), dim(Xob)[1])  +
  d1$mean * Xob[,1,1] + 
  d2$mean * Xob[,1,2] +
  d3$mean * Xob[,1,3] 
logitp_v2 <- matrix(d0$mean, length(d0$mean), dim(Xob)[1])  +
  d1$mean * Xob[,2,1] + 
  d2$mean * Xob[,2,2] +
  d3$mean * Xob[,2,3] 

p_v1 <- exp(logitp_v1)/(exp(logitp_v1) + 1)
p_v2 <- exp(logitp_v2)/(exp(logitp_v2) + 1) 

# mean probability of detection, all species, all sites, on visit 1 and visit 2
mean(p_v1); mean(p_v2)

# plot probability of detection
tibble(
  mean_p_v1 = rowMeans(p_v1),
  mean_p_v2 = rowMeans(p_v2)
) %>% 
  arrange(-mean_p_v2, -mean_p_v1) %>% 
  add_column(species = 1:M) %>% 
  ggplot(., aes(x=species, ymin=mean_p_v1, ymax=mean_p_v2)) +
  geom_hline(yintercept=mean(p_v1), color="red", lty="dotted") +
  geom_hline(yintercept=mean(p_v2), color="red") +
  geom_linerange() +
  geom_point(aes(y=mean_p_v1), pch=1) +
  geom_point(aes(y=mean_p_v2), color="black") +
  labs(x="Species, sorted by probability of detection (visit 2)", 
       y="Mean probability of detection over all sites\nOpen circles = visit 1; Closed circles = visit 2")

```
