---
title: "bayesBirdModel_multiSPecies_v6 - FARM"
author: "Liz Law"
date: "22 Mar 2022"
output: html_document
---
workingconservation@gmail.com

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updates from v5:

* Discussed with Chloé Nater (NINA)
* simplified model even further:

* removed tau (not required in nimble)
* removed omega parameter completely (doesn't do anything here, as we don't select from a regional pool - we estimate directly the number of missing species via rarefication)
* assume that all npsi (occ parameters) share the same sd.betalpsi
* assume that all np (obs parameters) share the same sd.betalp

Other options not yet implemented here
… assume that some npsi (occ parameters) share the same mu.betalpsi
… assume that some np (obs parameters) share the same mu.betalp 
… assume that some species (k) share the same betalpsi and/or betalp

The most pragmatic of the latter might be to assume all species that are seen rarely over the entire study are given the same parameters. I've also added in species groups, partly in preperation for this

# Setup
```{r library setup}
library(nimble)
library(tidyverse)
myversion <- "v6"
```

# build model
```{r model code}
occupancy_code <- nimbleCode({
  
  # Priors
  
    # Proportion of species present in the study region assuming nSpecies > nObsSpecies
    # omega ~ dunif(0,1)  
    
    # Hyperpriors describe the community
  
    # For the model of occupancy (psi)
    mu.lpsi ~ dnorm(0,0.01)
    sd.lpsi ~ dunif(0, 12)   # bounds of uniform chosen by trial and error
    
    for (lpsii in 1:npsi){
      mu.betalpsi[lpsii] ~ dnorm(0, 2.25)          # recommended default priors
      #sd.betalpsi[lpsii]  ~ dunif(0, 6)
    }
    sd.betalpsi  ~ dunif(0, 6) # constant sd for all psi params

    # For the model of detection (p)
    mu.lp ~ dnorm(0, 0.1)
    sd.lp ~ dunif(0, 5) # bounds of uniform chosen by trial and error
    
    for (lpi in 1:np){
      mu.betalp[lpi] ~ dnorm(0, 2.25)
      # sd.betalp[lpi] ~ dunif(0, 5)
    }
    sd.betalp ~ dunif(0, 5) # constant sd for all p params
    
    # Priors for species-specific effects in occupancy and detection
    # are derived from the community level distributions
    for(k in 1:M){
      # occupancy
      lpsi[k] ~ dnorm(mu.lpsi, sd=sd.lpsi)    
      for (lpsii in 1:npsi){
         betalpsi[k, lpsii] ~ dnorm(mu.betalpsi[lpsii], sd=sd.betalpsi) #[lpsii])
      }
      
      # detection
      lp[k] ~ dnorm(mu.lp, sd=sd.lp)
      for(lpi in 1:np){
        betalp[k, lpi] ~ dnorm(mu.betalp[lpi], sd=sd.betalp) #[lpi])
      }
    }
    
    # Super-population process: proportion of species in the regional pool part of the community 
    #for(k in 1:M){
    #  w[k] ~ dbern(omega)  # w is binary presence of each species in the community
    #}
    
    # Ecological model for true occurrence (process model) - enviro variables
    for(k in 1:M){
      for (i in 1:nSite) {
        logit(psi[i,k]) <- lpsi[k] + 
           betalpsi[k,1] * Xoc[i,1] +  # elevation
           betalpsi[k,2] * Xoc[i,2] +  # fl_dis
           betalpsi[k,3] * Xoc[i,3]    # sidi1ha
          # betalpsi[k,4] * Xoc[i,4]   # slope 
          # betalpsi[k,5] * Xoc[i,5]   # farm_type !Binary
        mu.psi[i,k] <- psi[i,k] # * w[k]
        z[i,k] ~ dbern(mu.psi[i,k])
      }
    }
    
    # Observation model for replicated detection/non-detection observations 
    for(k in 1:M){
      for (i in 1:nSite){
        for(j in 1:nVisits){
          logit(p[i,j,k]) <- lp[k] + 
            betalp[k,1] * Xob[i,j,1] + # date
            betalp[k,2] * Xob[i,j,2] + # start1
            betalp[k,3] * Xob[i,j,3] #+ # start2
            # betalp[k,4] * Xob[i,j,4]   # visibility !Binary
          mu.p[i,j,k] <- z[i,k] * p[i,j,k]   
          Y[i,j,k] ~ dbern(mu.p[i,j,k])  
        }
      }
    }
    
    # Derived quantities
    for (i in 1:nSite){
      SR_site_all[i] <- sum(z[i,1:M])          # Number of occurring species at each site
      SR_site_obs[i] <- sum(z[i,1:Nobs])       # Number of observed species occurring species at each site
      SR_site_fspp[i] <- sum(z[i,1:Nobs] * fspp[1:Nobs])        # forest dependent species
      SR_site_mig[i] <- sum(z[i,1:Nobs] * mig[1:Nobs])          # migrants
      SR_site_fnDiet[i] <- sum(z[i,1:Nobs] * fnDiet[1:Nobs])    # fruit/nectar species
      SR_site_invDiet[i] <- sum(z[i,1:Nobs] * invDiet[1:Nobs])  # invertebrate species
    }
    #n0 <- sum(w[(Nobs+1):M])              # Number of unseen species over study
    #Ntotal <- sum(w[1:M])                 # Total metacommunity size
    
})
```

# Specify model: code, data (observed), constants (set), and inits (starting value for estimated values)

```{r import data}
nimData <- readRDS("/Users/elaw/Desktop/LeuphanaProject/BirdModelling/Leuphana_Bird_Modelling/nimbleMultiOcc/nimbleModel_multiOcc_v6_allData.RDS")

y_all <- nimData$y_all
farmSites <- nimData$farmSites
OccVarsFarm <- nimData$OccVarsFarm
ObsVarsFarm <- nimData$ObsVarsFarm
sppTraits <- nimData$sppTraits
Nadd <- 17 # based on individual rarefaction curves

y <- y_all[(dimnames(y_all)[[1]] %in% farmSites), , ]       # [site, visit, species]

nSite <- dim(y)[1]
nVisits <- dim(y)[2]
Nobs <- dim(y)[3]
M <- Nobs + Nadd
Y <- abind::abind(y, array(0, dim = c(dim(y)[1:2], Nadd)))  ## add the 'missing' species - not seen so = 0, 'absence'

# inits: if we observe them in a survey they are there
z <- array(apply(Y, c(1,3), sum, na.rm = TRUE), dim=dim(Y)[c(1,3)]); z[] <- ifelse(z > 0, 1, NA) 
# w <- c(rep(1, Nobs), rep(NA, Nadd)) 

Xoc <- OccVarsFarm %>% select(-point_id) %>% as.matrix() %>% array(dim = dim(.))   # [site, vars]
Xob <- array(NA, dim = c(nSite, nVisits, dim(ObsVarsFarm)[2]-2)) # [site, visit, vars]
Xob[,,1] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(date)
Xob[,,2] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(start1)
Xob[,,3] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(start2)
Xob[,,4] <- ObsVarsFarm %>% arrange(round, point_id) %>% pull(visibility)

# constrain to best functioning variables
Xoc <- Xoc[,1:3]
Xob <- Xob[,,1:3]

npsi <- dim(Xoc)[2]
np <- dim(Xob)[3]

# groups of interest
fspp <- sppTraits$forSpec 
mig <- sppTraits$migYes 
fnDiet <- sppTraits$FruiNect 
invDiet <- sppTraits$Invertebrate 

data <- list(Y = Y, 
             # v = v,
             z = z,
             #w = w,
             Xoc = Xoc,
             Xob = Xob)  
constants <- list(nSite = nSite, 
                  nVisits = nVisits, 
                  Nobs = Nobs,
                  Nadd =  Nadd, 
                  M = M,
                  npsi = npsi,
                  np = np,
                  fspp=fspp,
                  mig=mig,
                  fnDiet=fnDiet,
                  invDiet=invDiet)

# Initial values - match up with the number of observed vars used in the model
#omega <- Nobs/M
#wst <- rep(1, M)                   # Simply set everybody at occurring
zst <- array(1, dim = c(nSite, M)) # ditto
lpsi <- rnorm(n = M)
betalpsi <- array(rnorm(n = M * npsi), dim = c(M, npsi))
lp <- rnorm(n = M)
betalp <- array(rnorm(n = M * np), dim = c(M, np))

inits <- list(
  #omega = omega,
  z = zst, 
  #w = wst, 
  lpsi = lpsi, 
  betalpsi = betalpsi, 
  lp = rnorm(n = M), 
  betalp = betalp #,
  #Yb = Y
)

# Set 1
params1 <- c("mu.lpsi", "sd.lpsi", "mu.betalpsi", "sd.betalpsi", 
             "mu.lp", "sd.lp", "mu.betalp", "sd.betalp", 
             "SR_site_all", "SR_site_obs",
             "SR_site_fspp", "SR_site_mig","SR_site_fnDiet", "SR_site_invDiet") #,
             # "omega", "n0", "Ntotal")


```

## the short, default way
```{r run nimbleMCMC}

# MCMC settings - development - final
# (ni - nb)/nt # samples per chain
# ni <- 10      ;   nt <- 1    ;   nb <- 0       ;   nc <- 3
# ni <- 25000   ;   nt <- 10   ;   nb <- 0   ;   nc <- 3
 ni <- 50000   ;   nt <- 20   ;   nb <- 0   ;   nc <- 3
# ni <- 50000   ;   nt <- 20   ;   nb <- 15000   ;   nc <- 3
# ni <- 130000  ;   nt <- 30   ;   nb <- 25000   ;   nc <- 3   # = 3500 samples per chain

samples <- nimbleMCMC(
  code = occupancy_code,
  data = data,
  constants = constants,
  inits = inits,
  monitors = params1,
  thin = nt,
  niter = ni,
  nburnin = nb,
  nchains = nc)
```

## Or the long way
```{r not run manually compile build run}
# # Build the model
# occupancy_model <- nimbleModel(
#  occupancy_code,
#  data = data,
#  constants = constants,
#  inits = inits
# )
# 
# # configure the MCMC algorithm and build the MCMC object using default samplers
# MCMCconf <- configureMCMC(occupancy_model, monitors = params1)

# check samplers
# MCMCconf$printSamplers()

# we may want to change the samplers - testing automated blocking 
# https://r-nimble.org/html_manual/cha-mcmc.html#automated-parameter-blocking
# autoBlockConf <- configureMCMC(occupancy_model, monitors = params1, autoBlock = TRUE)
# 
# MCMC <- buildMCMC(autoBlockConf, monitors = params1, thin = nt)
# # 
# # # compile for faster execution
# CMCMC <- compileNimble(MCMC, project = occupancy_model)
# # 
# # ## Run the MCMC and extract the samples
# samples <- runMCMC(mcmc = CMCMC, niter = ni, nburnin = nb, nchains = nc)
```

```{r save results}
saveRDS(samples, paste0("BirdOccMod_samples_farm_", myversion, "_", format(Sys.time(), "%Y%m%d"), ".RDS"))
```
 
# model diagnostics
 
```{r MCMCvis}

library(MCMCvis)

pg1 <- c("SR_site_all") #"omega", "n0", "Ntotal", 
pg2 <-c("mu.lpsi", "sd.lpsi", "mu.betalpsi", "sd.betalpsi", "mu.lp", "sd.lp", "mu.betalp", "sd.betalp")

# summary of estimates
mySum1 <- MCMCsummary(samples, 
            params = pg1, 
            round = 2)

mySum2 <- MCMCsummary(samples, 
            params = pg2, 
            round = 2)

# trace plots
MCMCtrace(samples, 
          params = pg1,
          ind = TRUE, 
          Rhat = TRUE, n.eff = TRUE,
          filename = paste0("plots_traceDensity_pg1_farm_", myversion, "_", format(Sys.time(), "%Y%m%d"), ".pdf"))
MCMCtrace(samples, 
          params = pg2,
          ind = TRUE, 
          Rhat = TRUE, n.eff = TRUE,
          filename = paste0("plots_traceDensity_pg2_farm_", myversion, "_", format(Sys.time(), "%Y%m%d"), ".pdf"))

# caterpillar plots
MCMCplot(samples, 
         # samples for model 2,
         params = pg2, 
         ci = c(50, 90), 
         HPD = FALSE)  # plots equal tailed credible intervals

MCMCplot(samples, 
         # samples for model 2,
         params = 'SR_site_all', 
         ci = c(50, 90), 
         HPD = FALSE)  # plots equal tailed credible intervals
```

Model diagnostics of residuals
