---
title: "bayesBirdModel_multiSPecies_v5 - FARM"
author: "Liz Law"
date: "10 Jan 2022"
output: html_document
---
workingconservation@gmail.com

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updates from v4:

* Discussed with Chloé Nater (NINA)
* Remove missing outcome sites
* Species identified to genus as additional species (finalized version should compare removing these)
* Included only FARM and select variables with transformations for this.

# The model framework

Define, for each species:

* `y[i,j,k]` as the observation of a species `k` recorded from site `i` on visit `j`. `y` is binary.
* `psi[i,k]` is the occupancy probability
* `p[i,j,k]` is the detection probability
* `z[i,k]` is the latent state of whether a site is really occupied (binary) by species `k`
* `nSite` is the number of sites
* `nVisits` is the number of sampling visits to each site
* `Nobs` is the number of species observed in the study
* `M` is the total possible species pool, such that `M` >> `Nobs` if including unseen species, M = Nobs otherwise. 

*latent occupancy is linearly related to the environmental variables*
`z[i,k] ~ dbern(psi[i,k])`
`logit(psi[i,k]) = beta0[k] + beta1[k]*X1[i] + beta2[k]*X2[i] + ...` (plus other environmental predictors)

*observed occupancy is a function of occupancy and observability, the latter linearly related to spatial/temporal variables*
In this case, we assume only false negatives (not false positives)
`y[i,j,k] ~ dbern(z[i,k]*p[i,j,k])`
`logit(p[i,j,k]) = alpha0[k] + alpha1[k]*Y1[i,j] + alpha2[k]*Y2[i,j] + ...` (plus other environmental predictors)

*site specific species richness, and conditional species richness*
SR[i] = sum(psi[i,1:k])    # sum over k
SR_cond[i] = sum(z[i,1:k]) # sum over k for predictions conditional on the data observed at surveyed sites

*options*
Can either model as the collection of independent single-species occupancy models OR
link by modelling parameters as random effects (i.e. hyperparameters mu and sd to be estimated), and thereby leverage observations of more common species for estimation of less common ones:

`beta0[k] ~ dnorm(mu_b0, sd_b0)`
`beta1[k] ~ dnorm(mu_b1, sd_b1)`
.. and so on for all beta for all species k. (This assumes that the beta from all the species come from a normal distribution). 

Can include the species never observed in the study such that the total species is given by `M`. Note, while the choice of `M` is not critical, it should be >> the true number of species `N` (which is bigger than the observed). To do this, introduce `omega`, the probability the species is part of the community, resulting in the binary `w[k]`. This can then be summed to get the regional species richness pool. (we probably don't want this, particularly as it slows the computation considerably, and the interpretation of what N relates to is difficult).

`w[k] ~ dbern(omega)`
`z[i,k] ~ dbern(w[k]*psi[i,k])`
`N = sum(w[1:k])`  # sum over k

# Setup
```{r library setup}
library(nimble)
# library(nimbleEcology) - don't need this as not using a simple occupancy model
library(tidyverse)
```

# build model

Based on https://github.com/nimble-training/AHMnimble/blob/master/Chapter_11/Section_11p6p3_example_nimble.R
and https://github.com/nimble-training/AHMnimble/blob/master/Chapter_11/Section_11p7p2_example_nimble.R
and also check https://doi.org/10.1002/ece3.6053 

New updates to deal with missing data as advised by Chloé Nater.

```{r model code}
occupancy_code <- nimbleCode({
  
  # Priors
  
    # Proportion of species present in the study region assuming nSpecies > nObsSpecies
    omega ~ dbeta(0.001, 1)
    # option to truncate this to Nobs/M - 1 (can't be less than observed)
    # omega ~ dunif(Nobs/M,1)  
    # omega ~ T(dbeta(0.001, 1), Nobs/M, 1)  
    
    # Hyperpriors describe the community
  
    # For the model of occupancy (psi)
    mu.lpsi ~ dnorm(0,0.01)
    tau.lpsi <- pow(sd.lpsi, -2)  # power: pow(x,y) = x^y
    sd.lpsi ~ dunif(0, 12)   # bounds of uniform chosen by trial and error
    
    for (lpsii in 1:npsi){
      mu.betalpsi[lpsii] ~ dnorm(0, 2.25)          # recommended default priors
      tau.betalpsi[lpsii]  <- pow(sd.betalpsi[lpsii] , -2)
      sd.betalpsi[lpsii]  ~ dunif(0, 6)
    }

    # For the model of detection (p)
    mu.lp ~ dnorm(0, 0.1)
    tau.lp <- pow(sd.lp, -2)
    sd.lp ~ dunif(0, 5) # bounds of uniform chosen by trial and error
    
    for (lpi in 1:np){
      mu.betalp[lpi] ~ dnorm(0, 2.25)
      tau.betalp[lpi] <- pow(sd.betalp[lpi], -2)
      sd.betalp[lpi] ~ dunif(0, 5)
    }
    
    # Priors for species-specific effects in occupancy and detection
    # are derived from the community level distributions
    for(k in 1:M){
      # occupancy
      lpsi[k] ~ dnorm(mu.lpsi, tau.lpsi)    
      for (lpsii in 1:npsi){
         betalpsi[k, lpsii] ~ dnorm(mu.betalpsi[lpsii], tau.betalpsi[lpsii])
      }
      
      # detection
      lp[k] ~ dnorm(mu.lp, tau.lp)
      for(lpi in 1:np){
        betalp[k, lpi] ~ dnorm(mu.betalp[lpi], tau.betalp[lpi])
      }
    }
    
    # Super-population process: proportion of species in the regional pool part of the community 
    for(k in 1:M){
      w[k] ~ dbern(omega)  # w is binary presence of each species in the community
    }
    
    # Ecological model for true occurrence (process model) - enviro variables, interacting with habitat
    # here 1, 2, 3, represent: habitat, farm85, forest85, elevation, twi, dis, dis85
    for(k in 1:M){
      for (i in 1:nSite) {
        logit(psi[i,k]) <- lpsi[k] + 
          # series of betalpsi1[k] * X1[i]
          # inprod(betalpsi[k,1:npsi], Xoc[i,1:npsi])  
           betalpsi[k,1] * Xoc[i,1] +  # elevation
           betalpsi[k,2] * Xoc[i,2] +  # fl_dis
           betalpsi[k,3] * Xoc[i,3]    # sidi1ha
          # betalpsi[k,4] * Xoc[i,4]   # slope 
          # betalpsi[k,5] * Xoc[i,5]   # farm_type !Binary
        mu.psi[i,k] <- w[k] * psi[i,k]
        z[i,k] ~ dbern(mu.psi[i,k])
      }
    }
    
    # Observation model for replicated detection/non-detection observations 
    # n_observers + visibility + recording + date + start
    for(k in 1:M){
      for (i in 1:nSite){
        for(j in 1:nVisits){
          logit(p[i,j,k]) <- lp[k] + 
            # series of betalp1[k] * X1[i,j]
            # inprod(betalp[k,1:np], Xob[i,j,1:np]) 
            betalp[k,1] * Xob[i,j,1] + # date
            betalp[k,2] * Xob[i,j,2] + # start1
            betalp[k,3] * Xob[i,j,3] #+ # start2
            # betalp[k,4] * Xob[i,j,4]   # visibility !Binary
          mu.p[i,j,k] <- z[i,k] * p[i,j,k]   
          Y[i,j,k] ~ dbern(mu.p[i,j,k])  
        }
      }
    }
    
    # Derived quantities
    for (i in 1:nSite){
      SR_site_all[i] <- sum(z[i,1:M])          # Number of occurring species at each site
      SR_site_obs[i] <- sum(z[i,1:Nobs])       # Number of observed speciesoccurring species at each site
    }
    n0 <- sum(w[(Nobs+1):M])              # Number of unseen species over study
    Ntotal <- sum(w[1:M])                 # Total metacommunity size
    
})
```

# Specify model: code, data (observed), constants (set), and inits (starting value for estimated values)

```{r import data}
source("/Users/elaw/Desktop/LeuphanaProject/BirdModelling/Leuphana_Bird_Modelling/nimbleMultiOcc/nimbleModel_multiOcc_v5_farmData.R")
```

## the short, default way
```{r run nimbleMCMC}

# MCMC settings - development - final
# (ni - nb)/nt # samples per chain
# ni <- 10      ;   nt <- 1    ;   nb <- 0       ;   nc <- 3
# ni <- 25000   ;   nt <- 10   ;   nb <- 0   ;   nc <- 3
# ni <- 50000   ;   nt <- 20   ;   nb <- 0   ;   nc <- 3
# ni <- 50000   ;   nt <- 20   ;   nb <- 15000   ;   nc <- 3
 ni <- 130000  ;   nt <- 30   ;   nb <- 25000   ;   nc <- 3   # = 3500 samples per chain

samples <- nimbleMCMC(
  code = occupancy_code,
  data = data,
  constants = constants,
  inits = inits,
  monitors = params1,
  thin = nt,
  niter = ni,
  nburnin = nb,
  nchains = nc)
```

## Or the long way
```{r not run manually compile build run}
# # Build the model
# occupancy_model <- nimbleModel(
#  occupancy_code,
#  data = data,
#  constants = constants,
#  inits = inits
# )
# 
# # configure the MCMC algorithm and build the MCMC object using default samplers
# MCMCconf <- configureMCMC(occupancy_model, monitors = params1)

# check samplers
# MCMCconf$printSamplers()

# we may want to change the samplers - testing automated blocking 
# https://r-nimble.org/html_manual/cha-mcmc.html#automated-parameter-blocking
# autoBlockConf <- configureMCMC(occupancy_model, monitors = params1, autoBlock = TRUE)
# 
# MCMC <- buildMCMC(autoBlockConf, monitors = params1, thin = nt)
# # 
# # # compile for faster execution
# CMCMC <- compileNimble(MCMC, project = occupancy_model)
# # 
# # ## Run the MCMC and extract the samples
# samples <- runMCMC(mcmc = CMCMC, niter = ni, nburnin = nb, nchains = nc)
```

```{r save results}
saveRDS(samples, paste0("BirdOccMod_samples_farm", format(Sys.time(), "%Y%m%d"), ".RDS"))
```
 
# model diagnostics

```{r MCMCvis}

library(MCMCvis)

pg1 <- c("omega", "n0", "Ntotal", "SR_site_all", "SR_site_obs")
pg2 <-c("mu.lpsi", "sd.lpsi", "mu.betalpsi", "sd.betalpsi", "mu.lp", "sd.lp", "mu.betalp", "sd.betalp")

# summary of estimates
mySum1 <- MCMCsummary(samples, 
            params = pg1, 
            round = 2)

mySum2 <- MCMCsummary(samples, 
            params = pg2, 
            round = 2)

# trace plots
MCMCtrace(samples, 
          params = pg1,
          ind = TRUE, 
          Rhat = TRUE, n.eff = TRUE,
          filename = paste0("plots_traceDensity_pg1_farm", format(Sys.time(), "%Y%m%d"), ".pdf"))
MCMCtrace(samples, 
          params = pg2,
          ind = TRUE, 
          Rhat = TRUE, n.eff = TRUE,
          filename = paste0("plots_traceDensity_pg2_farm", format(Sys.time(), "%Y%m%d"), ".pdf"))

# caterpillar plots
MCMCplot(samples, 
         # samples for model 2,
         params = pg2, 
         ci = c(50, 90), 
         HPD = FALSE)  # plots equal tailed credible intervals

MCMCplot(samples, 
         # samples for model 2,
         params = 'SR_site_all', 
         ci = c(50, 90), 
         HPD = FALSE)  # plots equal tailed credible intervals
```
